\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-01-30 paper/main.tex
minor updates

\subsection{Evaluation and metrics}
As the number of drug-targets are sparsely given w.r.t. to the number of both drugs and proteins considered, the resulting training, validation and testing datasets are highly imbalanced. As there are only $~22.000$ links in the considered subset the ratio
\begin{equation*}
	\frac{\#dti\_links}{\#drugs \cdot \#proteins} \approx 360,
\end{equation*}

consequently needing compensation in the computed loss function and appropriate metrics for the evaluation.\\

On that account, we weighted all positive drug-protein pair samples with this ratio by introducing the following loss function w.r.t. to binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for given prediction $x$ and target $y$. We average this loss among all drug-protein pairs in the training set, leading to a stable environment for the used optimizing scheme \textit{Adam} \citep{Adam2014}. We implemented a 5-fold cross validation among the proteins as justified in the results section. Furthermore, we  used early stopping to detect plateaus in the training process. \\

To assess each model, we computed the \textit{area under receiver operating characteristic}-score (AUROC), \textit{F1-score} (F1) and \textit{Matthews correlation coefficient} (MCC) for each considered hyperparameter configuration setting.\\

Avoiding validation overfitting, we eventually tested the best performing model on the Yamanishi dataset as presented and discussed. Furthermore, we compare our results on this very dataset with various other cutting edge approaches to DTI-prediction.\\

\subsection{Model}
In order to build a method that incorporates both top-down and bottom-up features, we first created a model for each individually. Hence, we assemble rich molecular structure based features for drugs from \textit{SmilesTransformer}\citep{SmilesTransformer} and proteins from \textit{DeepGOPlus}\citep{DeepGoPlus}. \textit{SmilesTransformer} introduces an autoencoder, learning over the SMILES strings and thus the molecular organization of each drug in an unsupervised manner. On the other hand, we take advantage of the pretrained models of \textit{DeepGOPlus}, obtaining features from the proteins amino acid sequence, showing significant performance in the field of protein function prediction. Thus, both embeddings seem to suitably supplement the  following ontology based represenations.\\

In the top-down section, we used \textit{DL2vec}\citep{DL2vec2020} to obtain ontology based representations. Hereby, DL2vec constructs a graph by introducing vertices and edges for each ontology class and axiom, respectively, followed by random walks starting from each entity. These walks are eventually learned on using a Word2vec \citep{Word2vec2013} model. Thus, we pick up rich, neighbourhood focused representations for each entity, which has shown great results for representing protein function and phenotypes. The overall structure of the ontology can be seen in Figure~\ref{fig:Onto}.\\

@@

As we wanted to learn from the similarity of drug side effects and protein phenotypes we opted for a deep siamese network approach, hence learning a high-dimensional embedding emphasizing this identity by forcing a maximal cosine similarity between these embeddings. On the other hand we built a deep neural network for the molecular structure based features, also benefiting from the siamese network architecture. Therefore, the precomputed embeddings are run through a deep neural network transformation (DNNT) which also reduces the eventual representation size for drugs and proteins separately. An example structure for both types of features can be found in Figure~\ref{fig:SiameseNetwork}.\\

@@

\subsubsection{Graph convolutional layers}
These molecular and ontology based sub-models were added to a larger graph convolutional model, based on the protein-protein interaction (PPI) graph. The PPI dataset is represented by a graph $G=(V,E)$, where each protein is represented by a vertex $v\in V$, and each edge $e\in E\subseteq V\times V$ symbolizes an interaction between two proteins. Additionally, we introduce mapping $x:V\rightarrow\mathbb{R}^{d}$ projecting each vertex $v$ to its node feature $x_v = x(v)$, where $d$ denotes the dimensionality of the node features. \\

As described before, graph convolution has shown significant performance increase in a variety of tasks. While there are various methods out there we will only introduce the most basic one here. A graph convolutional layer w.r.t. \citet{GCNConv} hereby consists of a learnable weight matrix followed by an aggregation step, formalized by

\begin{equation}
	\mathbf{X}^{\prime} = \mathbf{\hat{D}}^{-1/2} \mathbf{\hat{A}}
	\mathbf{\hat{D}}^{-1/2} \mathbf{X} \mathbf{\Theta}
\end{equation}

where for a given graph $G=(V,E)$, $\hat{A} = A + I$ denotes the adjacency matrix with added self-loops for each vertex, $D$ is described by $\hat{D}_{ii} = \sum_{j=0} \hat{A}_{ij}$, a diagonal matrix displaying the degree of each node, and $\Theta$ denotes the learnable weight matrix. Added self-loops enforce that each node representation is directly dependent on its own preceding one. Notably, the number of graph convolutional layers stacked equals the radius of relevant nodes for each vertex within the graph. \\

The update rule for each individual node is denoted by

\begin{equation}
	\mathbf{x}^{\prime}_i = \mathbf{\Theta} \sum^{N}_{j}
	\frac{1}{\sqrt{\hat{d}_j \hat{d}_i}} \mathbf{x}_j
\end{equation}

where both $\hat{d}_i, \hat{d}_j$ are dependent on the edge weights $e_{ij}$ of the graph. With simple, single valued edge weights such as $e_{ij}=1 \text{ }\forall (i,j)\in E$, all $\hat{d}_i$ reduce to $d_i$, i.e. the degree of each vertex $i$. We will denote this type of graph convolutional neural layers with \textsc{GCNConv}. \\

While in this initial formulation the node-wise update step is defined by the sum over all neighbouring node representations, we are able to alter this formulation for a more sophisticated message passing scheme. As described we are able to rearrange the order of activation function $\sigma$, aggregation $\mathrm{AGG}$ and linear neural layer $\mathrm{MLP}$ with this formulation as proposed by \citet{GENConv2020}:

\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}

where we will generally only consider $\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We will denote this generalized layer type as \textsc{GENConv}, following the notation of \citet{PytorchGeometric}.
While the reordering is merely import for numerical stability, the authors claim that this alteration of the original formulation eases the vanishing gradient problem for deeper convolutional networks. Additionally, the authors revise the aggregation function by generalizing it, allowing more sophisticated weighting functions such as learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming signals for each vertex, substituting the previously mentioned averaging step. Hence, while \textsc{GCNConv} suffers from both vanishing gradients and signal fading for large scale, highly connected graphs, each propagation step in \textsc{GENConv} emphasizes strong signals with values close to $0$ and $1$. Note particularly, that the same convolutional filter and weight matrix are applied to and learned for all nodes simultaneously. Note additionally, that the resulting information hold no information on their own connectivity.\\

>>>>

\subsection{Evaluation and metrics}

\todo[inline]{First part is not about evaluation but optimization
  objectives, should go in the Model section:}
As the number of drug-targets are sparsely given w.r.t.\todo{no
  acronyms} to the number of both drugs and proteins considered, the
resulting training, validation and testing datasets are highly
imbalanced. As there are only $~22.000$\todo{exact number} links in
the considered subset the ratio
\begin{equation*}
	\frac{\#dti\_links}{\#drugs \cdot \#proteins} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function w.r.t.\todo{no
  acronym} to binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for given prediction $x$ and target $y$. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation among the
proteins as justified in the results section. Furthermore, we used
early stopping to detect plateaus in the training process.

\todo[inline]{This needs to be more elaborate: add the micro- vs
  macro-averages here (formulas)}
To assess each model, we computed the \textit{area under receiver
  operating characteristic}-score (AUROC), \textit{F1-score} (F1) and
\textit{Matthews correlation coefficient} (MCC) for each considered
hyperparameter configuration setting.

\todo[inline]{Remove from here:}
Avoiding validation overfitting, we eventually tested the best
performing model on the Yamanishi dataset as presented and
discussed. Furthermore, we compare our results on this very dataset
with various other cutting edge approaches to DTI-prediction.

\subsection{Model}

\todo[inline]{Introduce the notion of top-down and bottom-up:}
In order to build a method that incorporates both top-down and
bottom-up features, we first created a model for each
individually. Hence, we assemble rich molecular structure based
features for drugs from
\textit{SmilesTransformer} \citep{SmilesTransformer} and proteins from
\textit{DeepGOPlus} \citep{DeepGoPlus}. \textit{SmilesTransformer}
introduces an autoencoder, learning over the SMILES strings and thus
the molecular organization of each drug in an unsupervised manner. On
the other hand, we take advantage of the pretrained models of
\textit{DeepGOPlus}, obtaining features from the proteins amino acid
sequence, showing significant performance in the field of protein
function prediction. Thus, both embeddings seem to suitably supplement
the following ontology based representations.

In the top-down section, we used \textit{DL2vec} \citep{DL2vec2020} to
obtain ontology based representations. DL2vec constructs a graph by
introducing vertices and edges for each ontology class and axiom,
respectively, followed by random walks starting from each
entity. These walks are eventually learned on using a Word2vec
\citep{Word2vec2013} model. Thus, we pick up rich, neighbourhood
focused representations for each entity, which has shown great results
for representing protein function and phenotypes. The overall
structure of the ontology can be seen in Figure~\ref{fig:Onto}.

@@

As we wanted to learn from the similarity of drug side effects and
protein phenotypes we opted for a deep siamese network approach, hence
learning a high-dimensional embedding emphasizing this identity by
forcing a maximal cosine similarity between these embeddings. On the
other hand we built a deep neural network for the molecular structure
based features, also benefiting from the siamese network
architecture. Therefore, the precomputed embeddings are run through a
deep neural network transformation (DNNT) which also reduces the
eventual representation size for drugs and proteins separately. An
example structure for both types of features can be found in
Figure~\ref{fig:SiameseNetwork}.

@@

\subsubsection{Graph convolutional layers}
These molecular and ontology based sub-models were added to a larger
graph convolutional model, based on the protein-protein interaction
(PPI) graph. The PPI dataset is represented by a graph $G=(V,E)$,
where each protein is represented by a vertex $v\in V$, and each edge
$e\in E\subseteq V\times V$ symbolizes an interaction between two
proteins. Additionally, we introduce mapping
$x:V\rightarrow\mathbb{R}^{d}$ projecting each vertex $v$ to its node
feature $x_v = x(v)$, where $d$ denotes the dimensionality of the node
features.

As described before, graph convolution has shown significant
performance increase in a variety of tasks. While there are various
methods out there we will only introduce the most basic one here. A
graph convolutional layer \citet{GCNConv} consists of a
learnable weight matrix followed by an aggregation step, formalized by
\begin{equation}
	\mathbf{X}^{\prime} = \mathbf{\hat{D}}^{-1/2} \mathbf{\hat{A}}
	\mathbf{\hat{D}}^{-1/2} \mathbf{X} \mathbf{\Theta}
\end{equation}
where for a given graph $G=(V,E)$, $\hat{A} = A + I$ denotes the
adjacency matrix with added self-loops for each vertex, $D$ is
described by $\hat{D}_{ii} = \sum_{j=0} \hat{A}_{ij}$, a diagonal
matrix displaying the degree of each node, and $\Theta$ denotes the
learnable weight matrix. Added self-loops enforce that each node
representation is directly dependent on its own preceding
one. Notably, the number of graph convolutional layers stacked equals
the radius of relevant nodes for each vertex within the graph.

The update rule for each individual node is
\begin{equation}
	\mathbf{x}^{\prime}_i = \mathbf{\Theta} \sum^{N}_{j}
	\frac{1}{\sqrt{\hat{d}_j \hat{d}_i}} \mathbf{x}_j
\end{equation}
where both $\hat{d}_i, \hat{d}_j$ are dependent on the edge weights
$e_{ij}$ of the graph. With simple, single-valued edge weights such as
$e_{ij}=1 \text{ }\forall (i,j)\in E$, all $\hat{d}_i$ reduce to
$d_i$, i.e., the degree of each vertex $i$. We denote this type of
graph convolutional neural layers with \textsc{GCNConv}.

While in this initial formulation the node-wise update step is defined
by the sum over all neighbouring node representations, we are able to
alter this formulation to another message passing scheme.  We are able
to rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$ and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we will generally only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We will denote
this generalized layer type as \textsc{GENConv}, following the
notation of \citet{PytorchGeometric}.  While the reordering is mainly
import for numerical stability, this alteration also addresses the
vanishing gradient problem for deeper convolutional networks
\cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale, highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously, and the resulting information\todo{Which information?
  Specify} hold no information on their own connectivity.
